\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2016-11-22 review_response.Rnw

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}
\usepackage{amsmath}

%acronyms
% \acrodef{}{}

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\setlength{\parskip}{5mm}   
\setlength{\parindent}{0in}

\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

\begin{document}
\SweaveOpts{concordance=TRUE}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to reviewer comments ``NeuralNetTools: Visualization and Analysis Tools for Neural Networks'' by M. W. Beck.}

{\it I thank the reviewer for providing thoughtful comments on my manuscript and the NeuralNetTools package.  My response to these comments are shown in italics.  Page and paragraph numbers refer to the original manuscript.}

Reviewer C:

The manuscript identifies the need for open-source tools for visualizing the structure of a multilayer perceptron (MLP) neural network and for visualizing variable importance.  The author provides an intuitive background of neural networks and a brief description of previous publications that develop tools for visualizing and understanding the MLP neural network.  The supplied code works (mostly) as advertised (see below for discrepancies).  Overall the paper is well written with a logical flow.  The simple examples adequately demonstrate the tools.  However, in my opinion the applied example does not fully draw-out the additional interpretation that is claimed by the NeuralNetTools package.  Furthermore, there is limited usefulness in visualizing a neural network when there are more than about 10 input variables.  (Although the same problem applies to visualizing a tree-based model when there are many terminal nodes.)  Still I believe this tool could be useful for data sets with a small number of variables.

{\it I certainly agree that plotting a neural network with many variables is impractical, but as noted, this is an issue with any plotting method for large structures.  I believe the main benefit of the package is in the supporting functions, e.g., \texttt{garson}, \texttt{lekprofile}, \texttt{olden}, that use the structure of the model (weights) and predictions to better understand relationships between variables. The \texttt{plotnet} function has importance as a basic visualization tool that informs additional analysis. Our hope is that this is clear from the existing text in the manuscript that immediately follows the plot section (i.e., first paragraph, sec. 3.2).  I have added some text to the conclusions to re-emphasize this issue. First paraggraph of sec. 5: ``Although visualizing a neural network with \texttt{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.''

The applied example section has also been revised to provide a more comprehensive demonstration of the utility of the package.  Please see the specific response below.
}

I would recommend tentatively accepting the manuscript with the author addressing the following items:

$\bullet$ I would like to see a different applied example (or the current example changed) that gives a better practical demonstration of the ability to visualize a network and the corresponding interpretation.

{\it Content was added to the section that shows the interpretation of a neural network model using functions in the package.  Figure 7 was added to show the results from a simple model with an interpretation provided in the text.  Most of the addition to the text was included in the following paragraph (but see the applied example section for the entire revision):

``Figure 7 shows the information about arrival delays that can be obtained with the functions in \textbf{NeuralNetTools}. The NID (7a) shows the model structure and can be used develop a general characterization of the relationships between variables.  For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays. Figures 7b and 7c provide more quantitative descriptions using information from both the NID and model predictions.  Figure 7b shows variable importance using the \texttt{garson} and \texttt{olden} algorithms.  The \texttt{garson} function suggests time between destinations (\texttt{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \texttt{olden} method.  However, the \texttt{garson} function shows arrival time as the second most important variable, whereas the \texttt{olden} function shows this variable as having a weak negative relationship with arrival time. This discrepancy is explained below. Finally, results from the \texttt{lekprofile} function (7c) confirm those in 7b, with the addition of non-linear responses that vary by different groupings of the data.  Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group five that generally had early arrival times and long departure delays.''    
}

$\bullet$ Abstract:  please clarify what is meant by ``unstructured datasets.''

{\it This was changed to ``...among multiple variables''.}

$\bullet$ Introduction:  I do not agree that ``Data-intensive analysis is a relatively new research approach...''

{\it The first sentence was revised: ``A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data'.''}

$\bullet$ Page 2, last paragraph:  NeuralNetTools does not ``improve the breadth and quality of information obtained from the MLP neural network.''  Instead, the package provides tools to better understand the MLP neural network.

{\it The sentence was revised: ''...that was developed to better understand information obtained from the MLP neural network.''}

$\bullet$ Page 6:  I am not familiar with the concepts of power and degrees of freedom with neural networks.  Please clarify the sentence, ``Pruning has additional benefits...''

{\it  These sentences were clarified to emphasize that pruning can aid in 1) the interpretation of relationships between variables by removing unimportant connections, and 2) successive model fitting by reducing the number of weights that are changed during model training: ``In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.  This reduces the number of free parameters (weights) that are  estimated by the model optimization routine, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error (i.e., model identifiability, Ellenius and Groth 2000).'' }

$\bullet$ Page 8, Section 3.3:  It is not clear to me how the Lek profile is used to evaluate interactions between variables.

{\it The \texttt{lekprofile} function evaluates the change in the response variable across the range of values for a given explanatory variable while holding the remaining explanatory variables constant.  I have considered these plots conceptually similar to traditional interaction plots (i.e., what is the effect of an explanatory variable at different values of another explanatory variable), but I suppose there is not a 1:1 correspondence. The \texttt{lekprofile} function is more appropriately used to evaluate the shape of the response for a given explanatory variable, as opposed to the importance of variables as for the other functions. I've replaced `interactions' with `relationships' as this description is more appropriate.}

$\bullet$ Figure 6:  I was not able to reproduce this plot.  In addition, when I generated this plot, the bars for two of the groups contain zero (see Groups 1 and 2) - that is, the bar is greater than zero for part of the bar and is less than zero for the other part of the bar.   I thought these bars would be bounded above or below by zero—perhaps I'm not understanding this plot.

{\it This was an issue with stacking of non-positive values in ggplot bar plots.  The situation has been resolved.  Additionally, the default position of the bar plots is now \texttt{`dodge'} (side-by-side for each group), as passed to \texttt{ggplot2::geom\_bar}.  This was implemented in the development branch on the GitHub page and will be pushed in the next release to CRAN.}

$\bullet$ Section 4:  When I ran the code provided in the git repository, I could not reproduce Figure 7 (see attached Figure7.png file).  I added the following line before the toplo1 line to make this work:

\texttt{imp\_sums\$variable}

{\it Thanks, this was fixed.}

\end{document}

### 2016-11-22 nnt_manu.Rnw
revision draft done, need to proof and fix applied example

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural network packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, \code{plotnet}, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.  A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.  Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data. Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.  Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.

@@

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  Pruning has the additional benefit of increasing power by increasing the degrees of freedom or the identification of input nodes with low importance.  Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

<<eval = F, echo = F, results = hide>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

An alternative approach to evaluate variable interactions in a neural network is the Lek profile method \citep{Lek96,Gevrey03}. The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables. The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.

@@

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).  All explanatory variables were normalized and the response variable was standardized to 0--1.

The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.  Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

Several conclusions can be made from \cref{fig:flightimp}.  First, delays in arrival time are negatively related to distance and positively related to departure delays and air time.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
<<echo = F, results = hide>>=
# library("nnet")
# library("NeuralNetTools")
# library("dplyr")
# library("tidyr")
# library("ggplot2")
# library("nycflights13")
# 
# # data preprocessing
# # select UA carrier for month of december and relevant variables
# # normalize and standardize variables
# tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
#   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
#   mutate_each(funs(scale), -arr_delay) %>% 
#   mutate_each(funs(as.numeric), -arr_delay) %>% 
#   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
#   data.frame
# 
# # setup initial conditions
# nodes <- c(1, 5, 10)
# seeds <- sample(c(1:5000), 100)
# grids <- expand.grid(nodes, seeds)
# names(grids) <- c('nodes', 'seeds') 
# 
# # iterate through conditions
# imp_vals <- vector('list', nrow(grids))
# for(i in 1:nrow(grids)){
#   
#   cat(i, '\t')
#   
#   # get specific conditions
#   seed_val <- grids[i, 'seeds']
#   set.seed(seed_val)
#   node_val <- grids[i, 'nodes']
# 
#   # create model
#   mod <- nnet(arr_delay ~ ., size = node_val, linout = TRUE, data = tomod, trace = F)  
# 
#   # get importance, append to output
#   imp <- olden(mod, bar_plot = FALSE)
#   imp_vals[[i]] <- imp
#   
# }
# 
# # summarize
# imp_sums <- do.call('cbind', imp_vals)
# nms <- row.names(imp_sums)
# imp_sums <- t(imp_sums)
# names(imp_sums) <- nms
# row.names(imp_sums) <- 1:nrow(imp_sums)
# imp_sums <- data.frame(grids, imp_sums) %>% 
#   gather('variable', 'importance', -nodes, -seeds) %>% 
#   group_by(nodes, variable) %>% 
#   do((function(x) {
#        
#     out <- quantile(x$importance, c(0.05, 0.5, 0.95))
#     out <- data.frame(t(out))
#     names(out) <- c('lo', 'med', 'hi')
# 
#     out
#       
#   })(.)) %>% 
#   data.frame
# 
# save(imp_sums, file = 'imp_sums.RData')

@@

The \pkg{NeuralNetTools} package provides a simple approach to improve the quality of information obtained from a feed-forward \ac{mlp} neural network.  Functions can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).  Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods could be added based on popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).

>>>>

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns among multiple variables.  A common criticism of these methods is the inability to characterize relationships among variables from a fitted model. Although several techniques have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural network packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, \code{plotnet}, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.  Analysis is primarily focused on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.  Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data. Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.  Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.

@@

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to better understand information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.  This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}. Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

<<eval = T, echo = F, results = hide>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

An alternative approach to evaluate variable relationships in a neural network is the Lek profile method \citep{Lek96,Gevrey03}. The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables. The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.

@@

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate 1) use of the functions in \pkg{NeuralNetTools} to gain additional insight into relationships among variables, and 2) effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).

First, the apropriate month and airline carrier were selected, all explanatory variables were normalized, and the response variable was standardized to 0--1.
<<eval = T>>=
library("nycflights13")
library("dplyr")
tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
  mutate_each(funs(scale), -arr_delay) %>% 
  mutate_each(funs(as.numeric), -arr_delay) %>% 
  mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
  data.frame
@
Then, a standard \ac{mlp} with five hidden nodes was created with the \pkg{nnet} package to model the effects of selected variables on arrival delays. The entire dataset is used for the example but separate training and validation datasets should be used in practice.  
<<eval = F, eval = T, results = hide>>=
library(nnet)
mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
@
The default output is limited to structural information about the model and methods for model predictions (see \code{str(mod)} and \code{?predict.nnet}). Using functions from \pkg{NeuralNetTools}, a more comprehensive understanding of the relationships between the variables is illustrated.
<<eval = F>>=
plotnet(mod)
garson(mod)
olden(mod)
lekprofile(mod, group_vals = 5)
lekprofile(mod, group_vals = 5, group_show = TRUE)
@
<<echo = F, results = hide>>=
library(nnet)
mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)

pdf('figs/flightplots1.pdf', height = 4, width = 5.5, family = 'serif')
par(mar = c(0, 0, 0, 0))
plotnet(mod, cex_val = 0.7, line_stag = 0.02, circle_cex = 4)
gridExtra::grid.arrange(
  garson(mod),
  olden(mod),
  ncol = 1
  )
dev.off()

p1 <- lekprofile(mod, group_vals = 5)
p2 <- lekprofile(mod, group_vals = 5, group_show = TRUE)
pA <- ggplot_gtable(ggplot_build(p1))
pB <- ggplot_gtable(ggplot_build(p2))
maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
pA$widths[2:3] <- maxWidth
pB$widths[2:3] <- maxWidth

pdf('figs/flightplots2.pdf', height = 4.5, width = 8, family = 'serif')
gridExtra::grid.arrange(
  pA, 
  pB,
  ncol = 1
)
dev.off()
@

\Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}. The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used develop a general characterization of the relationships between variables.  For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays. \Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.  \Cref{fig:flightplotsb} shows variable importance using the \code{garson} and \code{olden} algorithms.  The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.  However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time. This discrepancy is explained below. Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.  Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.

%%%%
\begin{figure}
\centering
\subfloat[][\code{plotnet}]{
\includegraphics[width = 0.5\textwidth, page = 1]{figs/flightplots1.pdf}
\label{fig:flightplotsa}
}
\subfloat[][\code{garson} (top) and \code{olden} (bottom)]{
\includegraphics[width = 0.5\textwidth, page = 2]{figs/flightplots1.pdf}
\label{fig:flightplotsb}
}

\subfloat[][\code{lekprofile}]{
\includegraphics[width = 1.05\textwidth, page = 1]{figs/flightplots2.pdf}
\label{fig:flightplotsc}
}

\caption{Results from a simple \ac{mlp} model of arrival delay for December airline flights versus departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between destinations (\code{air\_time}), and distance flown (\code{distance}). The three plots show the \ac{nid} from \code{plotnet} (\ref{fig:flightplotsa}), variable importance with \code{garson} and \code{olden} (\ref{fig:flightplotsb}), and sensitivity analysis with variable groupings from \code{lekprofile} (\ref{fig:flightplotsc}). Interpretations are provided in the text.}
\label{fig:flightplots}
\end{figure}

A second analysis was needed to show the effects of network architecture and initial starting weights on uncertainty in estimates of variable importance. Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

Several conclusions from \cref{fig:flightimp} provide further information to interpret the trends in \cref{fig:flightplots}.  First, consistent relationships can be identified such that delays in arrival time were negatively related to distance and positively related to departure delays and air time.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
<<echo = F, results = hide>>=
# library("nnet")
# library("NeuralNetTools")
# library("dplyr")
# library("tidyr")
# library("ggplot2")
# library("nycflights13")
# 
# # data preprocessing
# # select UA carrier for month of december and relevant variables
# # normalize and standardize variables
# tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
#   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
#   mutate_each(funs(scale), -arr_delay) %>% 
#   mutate_each(funs(as.numeric), -arr_delay) %>% 
#   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
#   data.frame
# 
# # setup initial conditions
# nodes <- c(1, 5, 10)
# seeds <- sample(c(1:5000), 100)
# grids <- expand.grid(nodes, seeds)
# names(grids) <- c('nodes', 'seeds') 
# 
# # iterate through conditions
# imp_vals <- vector('list', nrow(grids))
# for(i in 1:nrow(grids)){
#   
#   cat(i, '\t')
#   
#   # get specific conditions
#   seed_val <- grids[i, 'seeds']
#   set.seed(seed_val)
#   node_val <- grids[i, 'nodes']
# 
#   # create model
#   mod <- nnet(arr_delay ~ ., size = node_val, linout = TRUE, data = tomod, trace = F)  
# 
#   # get importance, append to output
#   imp <- olden(mod, bar_plot = FALSE)
#   imp_vals[[i]] <- imp
#   
# }
# 
# # summarize
# imp_sums <- do.call('cbind', imp_vals)
# nms <- row.names(imp_sums)
# imp_sums <- t(imp_sums)
# names(imp_sums) <- nms
# row.names(imp_sums) <- 1:nrow(imp_sums)
# imp_sums <- data.frame(grids, imp_sums) %>% 
#   gather('variable', 'importance', -nodes, -seeds) %>% 
#   group_by(nodes, variable) %>% 
#   do((function(x) {
#        
#     out <- quantile(x$importance, c(0.05, 0.5, 0.95))
#     out <- data.frame(t(out))
#     names(out) <- c('lo', 'med', 'hi')
# 
#     out
#       
#   })(.)) %>% 
#   data.frame
# 
# save(imp_sums, file = 'imp_sums.RData')

@@

The \pkg{NeuralNetTools} package provides a simple approach to improve the quality of information obtained from a feed-forward \ac{mlp} neural network.  Functions can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).  Although visualizing a neural network with \code{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.  Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods could be added based on popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
